---
title: "GROUP WORK"
author: "ALPHA NUMERIC"
date: "2026-03-24"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["longtable", "fvextra"]
    keep_tex: true
fontsize: 10pt
header-includes:
  - \usepackage{microtype}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \sloppy
geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 60),  # wrapping long code lines
  comment = "#>",                       # cleaner output comments
  fig.align = "center"                  # center all plots
)

options(width = 60)   # wrap R *output* at ~60 chars

# Loading libraries
library(tidyverse)
library(transport)
library(lpSolve)
library(car)
library(knitr) # for kable tables
library(dplyr)
library(gridExtra)
library(ggplot2)
#library(ggfortify)
library(broom)
#library(ggcorrplot)
#library(corrr)
library(corrplot)
library(MASS)
library(datarium)
```

We proceeded to load marketing data from the 'datarium' library.\

```{r marketing data, echo=TRUE}
data("marketing", package = "datarium")
head(marketing)
```

Checking the names of the columns and some summary statistics.\

```{r overview, echo=TRUE}
names(marketing)
dim(marketing)
summary(marketing)
```

From the summary statistics, we were able to have a glimpse of our marketing dataset. It was evident that more budget was allocated to YouTube (176.45), than the rest of the marketing channels.\
We proceeded to investigate if our linear model assumptions were violated. We began with the most prominent one which is the **Assumption of Normality**.\

## Assumption of Normality checks

From our data, it was evident that our dependent variable was *sales* as our independent variables were *YouTube*, *Facebook* and *News Paper*.\
We began by use of a histogram. For the bin-width the range (Max - Min) from our summary statistics guided us.\

```{r YouTube, echo = TRUE}
ggplot(marketing, aes(x = youtube)) +
  geom_histogram(binwidth = 20, fill = "blue", color = "black") + labs(title = "Histogram of YouTube", x = "YouTube", y = "Count")
```

The YouTube Histogram appeared to be bi-modal.\

```{r Facebook, echo=TRUE}
ggplot(marketing, aes(x = facebook)) +
  geom_histogram(binwidth = 5, fill = "purple", color = "black") +
  labs(title = "Histogram of Facebook", x = "Facebook", y = "Count")
```

The histogram was uni-modal and appeared to be mesokurtic.\

```{r newspaper, echo=TRUE}
ggplot(marketing, aes(x = newspaper)) +
  geom_histogram(binwidth = 10, fill = "gold", color = "black") +
  labs(title = "Histogram of Newspaper", x = "Newspaper", y = "Count")
```

The histogram was positively skewed as it had a longer tail towards the right.\

```{r sales, echo=TRUE}
ggplot(marketing, aes(x = sales)) + geom_histogram(binwidth = 3, fill = "red", color = "black") +
  labs(title = "Histogram of Sales", x = "Sales", y = "Count")
```

The histogram appeared to be almost normally distributed.
